---
title: "Food at College"
output: 
  html_document:
    theme: cerulean
    code_folding: hide
---

```{r, warning=FALSE, message=FALSE}
# From your file menu (at the top of R-Studio) select:
# "Session -> Set working directory -> To source file location"
# Then play this chunk to get the data into R.
library(mosaic)
library(car)
library(DT)
library(pander)
library(readr)
food <- read_csv("food.csv") 
```


----

<!-- Instructions:

 Use the Food dataset and a Kruskal-Wallis Test(s) to answer an interesting question(s) that you come up with.
 
 About the dataset:
 
  * GPA - numerical, self-reported current college GPA

  
  * weight - numerical, self-reported weight in lbs


  * gender - "Female", "Male"


  * breakfast - "Cereal", "Donut" 
      The participants were shown a picture of "Cereal" and a "Donut" and asked which one of these pictures they associate with the word "breakfast."


  * calories_chicken - guessing calories in chicken piadina from the options of:
     1 - 265 
     2 - 430 
     3 - 610 
     4 - 720
    (the variable shows the actual number of calories participants selected)


   * calories_day - Importance of consuming calories per day
     1 - i dont know how many calories i should consume 
     2 - it is not at all important 
     3 - it is moderately important 
     4 - it is very important


   * comfort_food - List 3-5 comfort foods that come to mind. 
     Open ended 


   * comfort_food_reasons - What are some of the reasons that make you eat comfort food?      (i.e., anger, sadness, happiness, boredom, etc) - list up to three
     Open ended


   * comfort_food_reasons_coded
     1 - stress
     2 - boredom
     3 - depression/sadness
     4 - hunger
     5 - laziness
     6 - cold weather
     7 - happiness 
     8 - watching tv
     9 - none 


    * cook - how often do you cook?
     1 - Every day 
     2 - A couple of times a week 
     3 - Whenever I can, but that is not very often  
     4 - I only help a little during holidays 
     5 - Never, I really do not know my way around a kitchen


    * cuisine - what type of cuisine did you eat growing up?
     1 - American
     2 - Mexican.Spanish
     3 - Korean/Asian
     4 - Indian
     5 - American inspired international dishes
     6 - other


    * diet_current - describe your current diet
     open ended

	
    * diet_current_coded
     1 - healthy/balanced/moderated/
     2 - unhealthy/cheap/too much/random/
     3 - the same thing over and over
     4 - unclear


    * drink - which picture do you associate with the word “drink”?
     1 - orange juice
     2 - soda 


    * eating_changes  - Describe your eating changes since the moment you got into college?
     Open ended 


    * eating_changes_coded
     1 - worse
     2 - better
     3 - the same
     4 - unclear


    * eating_changes_coded1
     1 - eat faster
     2 - bigger quantity
     3 - worse quality 
     4 - same food
     5 - healthier
     6 - unclear
     7 - drink coffee 
     8 - less food
     9 - more sweets
     10 - timing 
     11 - more carbs or snacking
     12 - drink more water
     13 - more variety


    * eating_out - frequency of eating out in a typical week 
     1 - Never 
     2 - 1-2 times 
     3 - 2-3 times 
     4 - 3-5 times 
     5 - every day


    * employment - do you work? 
     1 - yes full time 
     2 - yes part time 
     3 - no
     4  - other


    * ethnic_food - How likely to eat ethnic food 
     1 - very unlikely 
     2 - unlikely 
     3 - neutral 
     4 - likely 
     5 - very likely 


    * exercise - how often do you exercise in a regular week?
     1 - Everyday 
     2 - Twice or three times per week 
     3 - Once a week
     4 - Sometimes 
     5 - Never


    * father_education - 
     1 - less than high school 
     2 - high school degree 
     3 - some college degree 
     4 - college degree 
     5 - graduate degree 


    * father_profession - what is your father profession?
     Open ended


    * fav_cuisine - What is your favorite cuisine?
     Open ended


    * fav_cuisine_coded
     0-none
     1 - Italian/French/greek
     2 - Spanish/mexican
     3 - Arabic/Turkish
     4 - asian/chineses/thai/nepal
     5 - American
     6 - African 
     7 - Jamaican
     8 - indian


    * fav_food - was your favorite food cooked at home or store bought? 
     1 - cooked at home 
     2 - store bought 
     3 - both bought at store and cooked at home


    * food_childhood - what was your favorite childhood food?
     Open ended



    * fries - which of these pictures you associate with word fries? 
     1 - Mcdonald’s fries
     2 - home fries

 
    * fruit_day - How likely to eat fruit in a regular day 
     1 - very unlikely 
     2 - unlikely 
     3 - neutral 
     4 - likely 
     5 - very likely 


    * grade_level – 
     1 - freshman 
     2 -Sophomore 
     3 - Junior 
     4 - Senior
     
      
    * greek_food - How likely to eat greek food when available?
     1 - very unlikely 
     2 - unlikely 
     3 - neutral 
     4 - likely 
     5 - very likely 
     
      
    * healthy_feel – how likely are you to agree with the following statement: “I feel very healthy!” ?
      1 to 10 where 1 is strongly agree and 10 is strongly disagree - scale
      
      
    * healthy_meal – what is a healthy meal? Describe in 2-3 sentences.
      Open ended
      
      
    * ideal_diet – describe your ideal diet in 2-3 sentences
      Open ended
      
      
    * Ideal_diet_coded
      1 – portion control
      2 – adding veggies/eating healthier food/adding fruit	
      3 – balance
      4 – less sugar
      5 – home cooked/organic
      6 – current diet
      7 – more protein
      8 – unclear
      
      
     * income
      1 - less than $15,000 
      2 - $15,001 to $30,000 
      3 - $30,001 to $50,000 
      4 - $50,001 to $70,000 
      5 - $70,001 to $100,000 
      6 - higher than $100,000
      
     * indian_food – how likely are you to eat indian food when available
      1 - very unlikely 
      2 - unlikely 
      3 - neutral 
      4 - likely 
      5 - very likely 
      
      
     * Italian_food – how likely are you to eat Italian food when available?
      1 - very unlikely 
      2 - unlikely 
      3 - neutral 
      4 - likely 
      5 - very likely 
      
      
     * life_rewarding – how likely are you to agree with the following statement: “I feel life is very rewarding!” ?
      1 to 10 where 1 is strongly agree and 10 is strongly disagree - scale
      
      
     * marital_status
      1 -Single 
      2 - In a relationship 
      3 - Cohabiting 
      4 - Married 
      5 - Divorced 
      6 - Widowed
      
      
     * meals_dinner_friend – What would you serve to a friend for dinner?
      Open ended
      
      
     * mothers_education 
      1 - less than high school 
      2 - high school degree 
      3 - some college degree 
      4 - college degree 
      5 - graduate degree
      
      
     * mothers_profession – what is your mother’s profession? 
      
      
     * nutritional_check - checking nutritional values frequency 
      1 - never 
      2 - on certain products only 
      3 - very rarely 
      4 - on most products 
      5 - on everything
      
      
     * on_off_campus – living situation
      1 - On campus 
      2 - Rent out of campus 
      3 - Live with my parents and commute 
      4 - Own my own house
      
      
     * parents_cook - Approximately how many days a week did your parents cook? 
      1 - Almost everyday 
      2 - 2-3 times a week 
      3 - 1-2 times a week 
      4 - on holidays only 
      5 - never
      
      
     * pay_meal_out - How much would you pay for meal out? 
      1 - up to $5.00 
      2 - $5.01 to $10.00 
      3 - $10.01 to $20.00 
      4 - $20.01 to $30.00 
      5 - $30.01 to $40.00 
      6 - more than $40.01 
      
      
     * Persian_food - How likely to eat persian food when available?
      1 - very unlikely 
      2 - unlikely 
      3 - neutral 
      4 - likely 
      5 - very likely 
      
      
     * self_perception_weight - self perception of weight 
      6 - i dont think myself in these terms 
      5 - overweight 
      4 - slightly overweight 
      3 - just right 
      2 - very fit 
      1 - slim 
      
      
     * Which of the two pictures you associate with the word soup?
      1 – veggie soup
      2 – creamy soup
      
      
     * sports - sports – do you do any sporting activity?
      1 - Yes 
      2 - No 
      99 – no answer
      
      
     * thai_food - How likely to eat thai food when available?
      1 - very unlikely 
      2 - unlikely 
      3 - neutral 
      4 - likely 
      5 - very likely 
      
      
     * tortilla_calories - guessing calories in a burrito sandwhich from Chipolte?
      1 - 580 
      2 - 725 
      3 - 940 
      4 - 1165
      
     * turkey_calories - Can you guess how many calories are in the foods shown below?
     (Panera Bread Roasted Turkey and Avocado BLT)
      1 - 345 
      2 - 500 
      3 - 690 
      4 - 850
      
      
     * type_sports – what type of sports are you involved?
      Open-ended
      
     * veggies_day - How likely to eat veggies in a day? 
      1 - very unlikely 
      2 - unlikely 
      3 - neutral 
      4- likely 
      5 - very likely
      
     * vitamins – do you take any supplements or vitamins?
      1 – yes
      2 – no
      
     * waffle_calories - guessing calories in waffle potato sandwhich 
      1 - 575 
      2 - 760 
      3 - 900 
      4 - 1315


Note this dataset and description come from: © 2020 Kaggle Inc for further details visit:

[Food choices on Kaggle](https://www.kaggle.com/borapajo/food-choices?select=food_coded.csv)


If you want to give your critiquers some ideas about what you have questions on and would like help with, place those details here.-->




<!--End your comments to your critiquers before this line.-->

----

<!--Begin writing your Analysis after this point.-->

## Background

This analysis uses the "Food Choices" dataset from Kaggle (© 2020 Kaggle Inc), which surveys college students about their eating habits, lifestyle choices, and demographic information. The dataset includes 125 students and contains various nutritional, dietary, and lifestyle variables.

After exploring the dataset, I became interested in whether students' exercise habits are associated with measurable health and academic outcomes. Exercise is often promoted as beneficial for both physical health and cognitive performance, but does this hold true in our college student population? This led me to examine two questions:

1. **Does exercise frequency relate to GPA?** Regular physical activity is said to improve focus, memory, and academic performance.

2. **Does exercise frequency relate to body weight?** Exercise is commonly associated with weight management and overall physical health.

To answer these questions, I will use the Kruskal-Wallis test, which is appropriate for comparing a quantitative variable (GPA or weight) across multiple groups (exercise frequencies) when we cannot assume the data are normally distributed.

----

## Analysis 1: Exercise Frequency and GPA

### Research Question

Is there a difference in median GPA among college students based on their exercise frequency?

### Hypotheses

**Null Hypothesis ($H_0$):** The median GPA is the same across all exercise frequency groups.

**Alternative Hypothesis ($H_a$):** At least one exercise frequency group has a different median GPA than the others.

**Significance Level:** $\alpha = 0.05$

### Analysis

```{r, warning=FALSE}
# Define labels for exercise frequency (excluding "Sometimes" and "Never")
exercise_labels <- c("1" = "Everyday", 
                     "2" = "2-3x a week", 
                     "3" = "Once a week")

# Clean data: remove missing values and filter for exercise groups 1-3
food$GPA <- as.numeric(food$GPA)
# Filter to keep only exercise levels 1, 2, and 3
food_subset <- subset(food, exercise %in% c(1, 2, 3))
gpa_exercise_clean <- na.omit(food_subset[, c("GPA", "exercise")])

# Create factor variable with meaningful labels
gpa_exercise_clean$exercise_factor <- factor(gpa_exercise_clean$exercise,
                                              levels = 1:3,
                                              labels = exercise_labels)

# Descriptive statistics
desc_stats_gpa <- favstats(GPA ~ exercise_factor, data = gpa_exercise_clean)

pander(desc_stats_gpa, 
       caption = "Descriptive Statistics: GPA by Exercise Frequency")
```

### Graphical Summary

```{r}
# Set larger bottom margin for rotated labels
par(mar = c(7, 4, 4, 2))

boxplot(GPA ~ exercise_factor, 
        data = gpa_exercise_clean,
        main = "Exercise Frequency Does Not Significantly Affect GPA",
        xlab = "",
        ylab = "GPA",
        col = "lightblue",
        border = "darkblue",
        las = 2,
        cex.axis = 0.75)

# Add x-axis label with more space
mtext("Exercise Frequency", side = 1, line = 5.5)

# Reset to default margins
par(mar = c(5, 4, 4, 2))
```

### Test Results

```{r}
# Perform Kruskal-Wallis test
kw_test_gpa <- kruskal.test(GPA ~ exercise_factor, data = gpa_exercise_clean)

# Pander the results
pander(kw_test_gpa)
```

### Interpretation

The Kruskal-Wallis test yielded a chi-squared statistic of `r round(kw_test_gpa$statistic, 3)` with `r kw_test_gpa$parameter` degrees of freedom and a p-value of `r round(kw_test_gpa$p.value, 4)`.

Since the p-value (`r round(kw_test_gpa$p.value, 4)`) is greater than our significance level of $\alpha = 0.05$, **we fail to reject the null hypothesis**.

**Conclusion:** There is no statistically significant evidence to suggest that median GPA differs based on exercise frequency among college students in this sample. The overall median GPA for these students is `r median(gpa_exercise_clean$GPA)`, and no group differed significantly from this value. The "Sometimes" and "Never" exercise groups were excluded from this analysis to focus on students with more defined exercise habits.

----

## Analysis 2: Exercise Frequency and Weight

### Research Question

Is there a difference in median body weight among college students based on their exercise frequency?

### Hypotheses

**Null Hypothesis ($H_0$):** The median weight is the same across all exercise frequency groups.

**Alternative Hypothesis ($H_a$):** At least one exercise frequency group has a different median weight than the others.

**Significance Level:** $\alpha = 0.05$

### Analysis

```{r, warning=FALSE}
# Clean data: remove missing values and filter for exercise groups 1-3
food$weight <- as.numeric(food$weight)
# Filter to keep only exercise levels 1, 2, and 3
food_subset_weight <- subset(food, exercise %in% c(1, 2, 3))
weight_exercise_clean <- na.omit(food_subset_weight[, c("weight", "exercise")])

# Create factor variable with meaningful labels
weight_exercise_clean$exercise_factor <- factor(weight_exercise_clean$exercise,
                                                 levels = 1:3,
                                                 labels = exercise_labels)

# Descriptive statistics
desc_stats_weight <- favstats(weight ~ exercise_factor, data = weight_exercise_clean)

pander(desc_stats_weight, 
       caption = "Descriptive Statistics: Weight (lbs) by Exercise Frequency")
```

### Graphical Summary

```{r}
# Set larger bottom margin for rotated labels
par(mar = c(7, 4, 4, 2))

boxplot(weight ~ exercise_factor, 
        data = weight_exercise_clean,
        main = "Exercise Frequency Does Not Significantly Affect Weight",
        xlab = "",
        ylab = "Weight (lbs)",
        col = "lightgreen",
        border = "darkgreen",
        las = 2,
        cex.axis = 0.75)

# Add x-axis label with more space
mtext("Exercise Frequency", side = 1, line = 5.5)

# Reset to default margins
par(mar = c(5, 4, 4, 2))
```

### Test Results

```{r}
# Perform Kruskal-Wallis test
kw_test_weight <- kruskal.test(weight ~ exercise_factor, data = weight_exercise_clean)

# Pander the results
pander(kw_test_weight)
```

### Interpretation

The Kruskal-Wallis test produced a chi-squared statistic of `r round(kw_test_weight$statistic, 3)` with `r kw_test_weight$parameter` degrees of freedom and a p-value of `r round(kw_test_weight$p.value, 4)`.

Since the p-value (`r round(kw_test_weight$p.value, 4)`) is `r ifelse(kw_test_weight$p.value < 0.05, "less", "greater")` than our significance level of $\alpha = 0.05$, **we `r ifelse(kw_test_weight$p.value < 0.05, "reject", "fail to reject")` the null hypothesis**.

**Conclusion:** `r ifelse(kw_test_weight$p.value < 0.05, "There is statistically significant evidence to suggest that median weight differs based on exercise frequency among college students in this sample.", paste("There is no statistically significant evidence to suggest that median weight differs based on exercise frequency among college students in this sample. The overall median weight for these students is", median(weight_exercise_clean$weight), "lbs, and no group differed significantly from this value."))`










